---
title: "Trend Following System - Apple Stock"
description: |
  Calculate slow and fast exponential moving averages for AAPL stock using historical data, visualize results and calculate return and performance metrics. Our strategy purchases the stock as the fast  exponential moving average crosses above the slow moving average. This system does not go short.
author: 
  - name: Alex Labuda
    affiliation: Suny New Paltz School of Business - Analytics Capstone
  - name: Camila Campi
  - name: Deniz Ahman
date: "9/9/2021"
output:
  distill::distill_article:
    toc: true
    toc_depth: 4
---

```{r setup, echo=FALSE, include=FALSE}
# Import relevant packages for this script
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyquant)
library(ggplot2)
library(tidyverse)
library(plotly)
library(quantmod)
library(dplyr)
library(hrbrthemes)
```

# Retrieve Apple and S&P500 Historical Data

```{r echo=FALSE}
# silence warnings
getSymbols.warning4.0=FALSE

# stock data for Apple and the S&P 500
getSymbols("AAPL", src='yahoo', )
getSymbols("^GSPC", src='yahoo',)
getSymbols("DAL", src = 'yahoo',)
```

```{r ,echo=FALSE}
# Save apple data to a dataframe
df <- data.frame(Date=index(AAPL),coredata(AAPL))
# take most recent observations
df <- tail(df, 4000)
# reset index
rownames(df) = NULL


# Save DAL data to a dataframe
dal_df <- data.frame(Date=index(DAL),coredata(DAL))
# take most recent observations
dal <- tail(dal_df, 4000)
# reset index
rownames(dal) = NULL


# SPX to a dataframe
SPX <- data.frame(Date=index(GSPC),coredata(GSPC))
# take most recent observations
SPX <- tail(SPX, 4000)
# reset index
rownames(SPX) = NULL
```

```{r ,echo=FALSE}
# copy the data
raw_data = df
dal_df = dal
```

```{r,echo=FALSE}
# rename AAPL columns for simplicity
names(raw_data)[names(raw_data)=="AAPL.Open"] = "Open"
names(raw_data)[names(raw_data)=="AAPL.High"] = "High"
names(raw_data)[names(raw_data)=="AAPL.Low"] = "Low"
names(raw_data)[names(raw_data)=="AAPL.Close"] = "Close"
names(raw_data)[names(raw_data)=="AAPL.Volume"] = "Volume"
names(raw_data)[names(raw_data)=="AAPL.Adjusted"] = "Price"

# rename DAL columns for simplicity
names(dal_df)[names(dal_df)=="DAL.Open"] = "Open"
names(dal_df)[names(dal_df)=="DAL.High"] = "High"
names(dal_df)[names(dal_df)=="DAL.Low"] = "Low"
names(dal_df)[names(dal_df)=="DAL.Close"] = "Close"
names(dal_df)[names(dal_df)=="DAL.Volume"] = "Volume"
names(dal_df)[names(dal_df)=="DAL.Adjusted"] = "Price"

# rename SPX columns for simplicity
names(SPX)[names(SPX)=="GSPC.Open"] = "Open"
names(SPX)[names(SPX)=="GSPC.High"] = "High"
names(SPX)[names(SPX)=="GSPC.Low"] = "Low"
names(SPX)[names(SPX)=="GSPC.Close"] = "Close"
names(SPX)[names(SPX)=="GSPC.Volume"] = "Volume"
names(SPX)[names(SPX)=="GSPC.Adjusted"] = "Price"
```

## Sample Period

**Apple**
```{r,echo=FALSE}
# retrieve max and min dates
date_max = max(raw_data$Date)
date_min = min(raw_data$Date)

print(paste("Our sample considers OHLC data from", date_min, "through", date_max, "for Apple, Delta and the S&P500"))

summary(raw_data[,1:5], digits = 3)
sd(raw_data$Close)
```
**SPX**
```{r}
summary(SPX[,1:5])
sd(SPX$Close)
```
**Delta**
```{r}
summary(dal_df[,1:5])
sd(dal_df$Close)
```


## Calculate Daily Return

Calculate a simple day-to-day return, from adjusted-close to adjusted-close and store in `return` and `log_return` vector:

```{r,echo=FALSE}
# create a blank column
raw_data$return = NA
raw_data$log_return = NA

# loop through data and calculate daily returns
for(t in 2:nrow(raw_data)){
  raw_data$return[t] = (raw_data$Price[t] - raw_data$Price[t-1])/ raw_data$Price[t-1]
  raw_data$log_return[t] = log(raw_data$Price[t]/raw_data$Price[t-1])
}


# Do the same for DAL
dal_df$return = NA
dal_df$log_return = NA

for(t in 2:nrow(dal_df)){
  dal_df$return[t] = (dal_df$Price[t] - dal_df$Price[t-1])/ dal_df$Price[t-1]
  dal_df$log_return[t] = log(dal_df$Price[t]/dal_df$Price[t-1])
}


# Do the same for SP500
SPX$return = NA
SPX$log_return = NA

for(t in 2:nrow(SPX)){
  SPX$return[t] = (SPX$Price[t] - SPX$Price[t-1])/ SPX$Price[t-1]
  SPX$log_return[t] = log(SPX$Price[t]/SPX$Price[t-1])
}


head(subset(raw_data, select = c("Date", "Close", "return", "log_return")))
head(subset(dal_df, select = c("Date", "Close", "return", "log_return")))
head(subset(SPX, select = c("Date", "Close", "return", "log_return")))
```

# Indicators and Studies

## Simple Moving Averages Functions

Create a function to calculate two simple moving averages of differing lengths for use for a trend-following trading system *(for this system we won't be using a simple moving average)*:

```{r ,echo=FALSE}
Two_MovAvg_function = function(variable, slow_period = 100, fast_period = 20){
  # blank list of variables
  fast_ma = rep(NA, length(variable))
  slow_ma = rep(NA, length(variable))
  
  # loop to calculate average price for each record
  for (t in (slow_period+1):length(variable)){
    est_slow = mean(variable[(t-(slow_period+1)):(t-1)])
    est_fast = mean(variable[(t-(fast_period+1)):(t-1)])
    # update current average price
    fast_ma[t] = est_fast
    slow_ma[t] = est_slow
  }
  # create a dataframe for new vectors
  ma_data = data.frame(fast_ma = fast_ma,
                       slow_ma  = slow_ma)
  return(ma_data)
}
```

## Exponential Moving Averages Function

Create a function to calculate two exponential moving averages of differing lengths for use for a trend-following trading system  and store in `fast_ma`, `slow_ma` and `ema_diff` vector:

```{r ,echo=FALSE}
EA_Mov_Avg = function(variable, slow_lag = 100, fast_lag = 25){
  fast_ea = variable
  slow_ea = variable
  ema_diff = rep(NA, length(variable))
  
  # Loop to calculate each exponential average
  for (t in 2:length(variable)){ 
      est_slow = slow_ea[t-1] + (slow_ea[t] - slow_ea[t-1]) / ((slow_lag + 1) / 2)
      est_fast = fast_ea[t-1] + (fast_ea[t] - fast_ea[t-1]) / ((fast_lag + 1) / 2)
      
      ema_diff[t] = est_slow - est_fast
      
    slow_ea[t] = est_slow
    fast_ea[t] = est_fast
  }
  # dataframe for new vector
  ema_data = data.frame(fast_ema = fast_ea,
                        slow_ema = slow_ea,
                        ema_diff = ema_diff)
  return(ema_data)
}
```

```{r ,echo=FALSE}
# add these as new vectors to original data 

new_df = EA_Mov_Avg(raw_data$Close, slow_lag = 150, fast_lag = 25)
new_dal_df = EA_Mov_Avg(dal_df$Close, slow_lag = 150, fast_lag = 25)

# Apple
raw_data$slow_ma = new_df$slow_ema
raw_data$fast_ma = new_df$fast_ema
raw_data$ema_diff = new_df$ema_diff

# DAL
dal_df$slow_ma = new_dal_df$slow_ema
dal_df$fast_ma = new_dal_df$fast_ema
dal_df$ema_diff = new_dal_df$ema_diff


tail(raw_data)
tail(dal_df)
```

## Average True Range

-   **The average true range (ATR)** is a technical analysis indicator, introduced by market technician J. Welles Wilder Jr. in his book New Concepts in Technical Trading Systems, that [measures market volatility](https://www.investopedia.com/articles/trading/08/average-true-range.asp) by decomposing the entire range of an asset price for that period.
-   The true range indicator is taken as the greatest of the following: current high less the current low; the [absolute value](https://www.investopedia.com/terms/a/absolute-value.asp) of the current high less the previous close; and the absolute value of the current low less the previous close. The ATR is then a [moving average](https://www.investopedia.com/terms/m/movingaverage.asp), generally using 14 days, of the true ranges.
-   **Here we can use this as a measure of recent volatility** in order to adjust our trade entry position size accordingly. The idea is that we can enter with smaller position sizes during times of greater volatility and larger positions during times of less volatility in an attempt to decrease  risk.

### True Range

First calculate the maximum of the aforementioned ranges and store in `max_range` vector:

```{r, echo=FALSE}
max_range_functon = function(variable){
  max_range = rep(0, length(variable))
  
  for (t in 2:length(variable$Price)){
    max_rng = max(abs(variable$High[t] - variable$Low[t]), abs(variable$High[t] - variable$Close[t-1]), abs(variable$Close[t-1] - variable$Low[t]))
    
    max_range[t] = max_rng
  }
  max_range_data = data.frame(max_range = max_range)
  
  return(max_range_data)
}

# Test Function


# Implement on full APPL dataset
max_range_df = max_range_functon(raw_data)
# Implement on full DAL dataset
max_range_dal_df = max_range_functon(dal_df)


# Create new variable in data
raw_data$max_range = max_range_df$max_range
dal_df$max_range = max_range_dal_df$max_range

subset(raw_data[25:32,], select = c("Date", "Open", "High", "Low", "Close", "max_range"))
subset(dal_df[25:32,], select = c("Date", "Open", "High", "Low", "Close", "max_range"))
```

### Average True Range Function

Create function that uses the true range vector to calculate a moving average of the true range with a user-specified lag period and store in `atr` vector:

```{r, echo=FALSE}
ATR_function = function(variable, lag = 15){
  atr_list = rep(NA, length(variable))
  
  # Loop to create ATR
  for (t in (lag+2):length(variable)){
    est_atr = mean(variable[(t-(lag)):t])
    
    atr_list[t] = est_atr
  }
  
  atr_data = data.frame(atr = atr_list)
  return(atr_data)
}
# Add vector to our data
## AAPL
atr_data = ATR_function(raw_data$max_range)
raw_data$atr = atr_data$atr

## DAL
atr_dal_data = ATR_function(dal_df$max_range)
dal_df$atr = atr_dal_data$atr
```

### Risk per lot

This is a multiple of the ATR for volatility-based position sizing:

Store in `risk_per_lot` vector:

```{r, echo=FALSE}
risk_multiplier = 5
equity = 100000


raw_data$risk_per_lot = raw_data$atr * risk_multiplier
subset(raw_data[200:210,], select = c("Close", "max_range", "atr", "risk_per_lot")) 

dal_df$risk_per_lot = dal_df$atr * risk_multiplier
subset(dal_df[200:210,], select = c("Close", "max_range", "atr", "risk_per_lot")) 
```

```{r, echo=FALSE}
# save moving average cross-over dates for chart annotations

h = raw_data[which((raw_data$fast_ma > raw_data$slow_ma) & (raw_data$fast_ma[-1] < raw_data$slow_ma[-1]))+1,]
l = raw_data[which((raw_data$fast_ma < raw_data$slow_ma) & (raw_data$fast_ma[-1] > raw_data$slow_ma[-1]))+1,]
h = na.exclude(h)
h
l


# save moving average cross-over dates for chart annotations

hi = dal_df[which((dal_df$fast_ma > dal_df$slow_ma) & (dal_df$fast_ma[-1] < dal_df$slow_ma[-1]))+1,]
lo = dal_df[which((dal_df$fast_ma < dal_df$slow_ma) & (dal_df$fast_ma[-1] > dal_df$slow_ma[-1]))+1,]
hi = na.exclude(hi)
hi
lo = na.exclude(lo)
lo
```

```{r, echo=FALSE}
# Create arrow annotations for trade entry signal

# AAPL
# Upper annotations
h_a <- list(
  x = h$Date,
  y = h$slow_ma,
  text = "Sell",
  xref = "x",
  yref = "y",
  showarrow = TRUE,
  arrowcolor = "red",
  arrowhead = 5,
  arrowsize = 0.8,
  arrowwidth = 1.5,
  opacity = 0.75,
  align = "left",
  ax = 5,
  ay = -55
)

# Lower annotations

l_a <- list(
  x = l$Date,
  y = l$slow_ma,
  text = "Buy",
  xref = "x",
  yref = "y",
  showarrow = TRUE,
  arrowcolor = "green",
  arrowhead = 5,
  arrowsize = 0.8,
  arrowwidth = 1.5,
  opacity = 0.75,
  align = "right",
  ax = -5,
  ay = 55
)


# figure labels
f <- list(
  family = "Courier New, monospace",
  size = 18,
  color = "#7f7f7f ")
x <- list(
  title = "x Axis",
  titlefont = f)
y <- list(
  title = "y Axis",
  titlefont = f)


# DAL

# Upper annotations
h_a_dal <- list(
  x = hi$Date,
  y = hi$slow_ma,
  text = "Sell",
  xref = "x",
  yref = "y",
  showarrow = TRUE,
  arrowcolor = "red",
  arrowhead = 5,
  arrowsize = 0.8,
  arrowwidth = 1.5,
  opacity = 0.75,
  align = "left",
  ax = 5,
  ay = -55
)

# Lower annotations
l_a_dal <- list(
  x = lo$Date,
  y = lo$slow_ma,
  text = "Buy",
  xref = "x",
  yref = "y",
  showarrow = TRUE,
  arrowcolor = "green",
  arrowhead = 5,
  arrowsize = 0.8,
  arrowwidth = 1.5,
  opacity = 0.75,
  align = "right",
  ax = -5,
  ay = 55
)


# figure labels
f <- list(
  family = "Courier New, monospace",
  size = 18,
  color = "#7f7f7f ")
x <- list(
  title = "x Axis",
  titlefont = f)
y <- list(
  title = "y Axis",
  titlefont = f)
```

# **Candlestick Chart**

## Visualizing the Trend-following System for AAPL

-   The following candlestick chart helps us to visualize what our strategy is doing

-   As the faster price smoothing crosses above the slower lag, the system will buy on that day.


```{r, echo=FALSE}
raw_data$MACD_direction = rep(NA, length(raw_data$Date))


# colors column for increasing and decreasing
for (i in 2:length(raw_data[,1])) {
  if (raw_data$ema_diff[i] > 0) {
      raw_data$MACD_direction[i] = 'Increasing'
  } else {
      raw_data$MACD_direction[i] = 'Decreasing'
  }
}

# color of bars for chart
i <- list(line = list(color = '#17BECF')) # 'i' for increasing
d <- list(line = list(color = '#b22222')) # 'd' for decreasing


# Plot main data
fig <- raw_data %>% plot_ly(x = ~Date, type="ohlc",
          open = ~Open, close = ~Close,
          high = ~High, low = ~Low, name = "AAPL",
          increasing = i, decreasing = d)

# Add Fast and Slow moving average lines
fig <- fig %>% add_lines(x = ~Date, y = ~slow_ma, name = "Slow",
            line = list(color = '#ccc', width = 0.5),
            legendgroup = "Bands", inherit = F,
            showlegend = TRUE, hoverinfo = "none") 
fig <- fig %>% add_lines(x = ~Date, y = ~fast_ma, name = "Fast",
            line = list(color = '#E377C2', width = 0.5),
            hoverinfo = "none", inherit = F)
# Add y-axis title
fig <- fig %>% layout(yaxis = list(title = "Price"))


# plot MACD line subplot
fig2 <- raw_data 
fig2 <- fig2 %>% plot_ly(x=~Date, y=~ema_diff, type='bar', name = "MACD",
                         color = ~MACD_direction, colors = c("red", "green"), alpha = 0.8)
fig2 <- fig2 %>% layout(yaxis = list(title = "MACD"))

# use subplot to add plots to the same figure
fig = subplot(fig, fig2, heights = c(0.8,0.15), nrows = 2,
              shareX = TRUE, titleY = TRUE)

# Add arrow annotations
fig <- fig %>% layout(annotations = h_a)
fig <- fig %>% layout(annotations = l_a)

# Add title
fig <- fig %>% layout(title = "<b> AAPL - Trend Following System </b>",
                       xaxis = list(rangeslider = list(visible = F)))

fig
```

## Visualizing the Trend-following System for DAL

- Here you can see that our system has a difficultly trading securities without clearly identifiable trends.
- Delta Airlines (DAL) does not have as clear of a bullish trend as we can see with Apple's stock. This is a pitfall for this type of trend-following system.

```{r, echo=FALSE}
dal_df$MACD_direction = rep(NA, length(dal_df$Date))


# colors column for increasing and decreasing
for (i in 2:length(dal_df[,1])) {
  if (dal_df$ema_diff[i] > 0) {
      dal_df$MACD_direction[i] = 'Increasing'
  } else {
      dal_df$MACD_direction[i] = 'Decreasing'
  }
}

# color of bars for chart
i <- list(line = list(color = '#17BECF')) # 'i' for increasing
d <- list(line = list(color = '#b22222')) # 'd' for decreasing


# Plot main data
fig3 <- dal_df %>% plot_ly(x = ~Date, type="ohlc",
          open = ~Open, close = ~Close,
          high = ~High, low = ~Low, name = "DAL",
          increasing = i, decreasing = d)

# Add Fast and Slow moving average lines
fig3 <- fig3 %>% add_lines(x = ~Date, y = ~slow_ma, name = "Slow",
            line = list(color = '#ccc', width = 0.5),
            legendgroup = "Bands", inherit = F,
            showlegend = TRUE, hoverinfo = "none") 
fig3 <- fig3 %>% add_lines(x = ~Date, y = ~fast_ma, name = "Fast",
            line = list(color = '#E377C2', width = 0.5),
            hoverinfo = "none", inherit = F)
# Add y-axis title
fig3 <- fig3 %>% layout(yaxis = list(title = "Price"))


# plot MACD line subplot
fig4 <- dal_df 
fig4 <- fig4 %>% plot_ly(x=~Date, y=~ema_diff, type='bar', name = "MACD",
                         color = ~MACD_direction, colors = c("red", "green"), alpha = 0.8)
fig4 <- fig4 %>% layout(yaxis = list(title = "MACD"))

# use subplot to add plots to the same figure
fig3 = subplot(fig3, fig4, heights = c(0.8,0.15), nrows = 2,
              shareX = TRUE, titleY = TRUE)

# Add arrow annotations
fig3 <- fig3 %>% layout(annotations = h_a_dal)
fig3 <- fig3 %>% layout(annotations = l_a_dal)

# Add title
fig3 <- fig3 %>% layout(title = "<b> DAL Stock - Moving Average Crossover </b>",
                       xaxis = list(rangeslider = list(visible = F)))

fig3
```

# Generate signals using MA Indicator

## Apple Stock

Now we use our indicator to generate trading signals and store them in `signal` vector:

```{r, echo=FALSE}
raw_data$signal = NA

# Loop through data to create signal column
for(t in 100:nrow(raw_data)-1){
  
  current_slow = raw_data$slow_ma[t]
  current_fast = raw_data$fast_ma[t]
  
  prev_slow = raw_data$slow_ma[t-1]
  prev_fast = raw_data$fast_ma[t-1]
  
  # use 'if' statement to translate crossover into signals
  if(current_fast > current_slow & prev_fast < prev_slow) {signal = "B"} else 
    if(current_fast < current_slow & prev_fast > prev_slow) {signal = "S"} else
      {signal = "H"}
  
  raw_data$signal[t+1] = signal
}

subset(raw_data[200:210,], select = c("Date","Close", "slow_ma", "fast_ma", "signal"))
```

## DAL Stock

```{r, echo=FALSE}
dal_df$signal = NA

# Loop through data to create signal column
for(t in 100:nrow(dal_df)-1){
  
  current_slow = dal_df$slow_ma[t]
  current_fast = dal_df$fast_ma[t]
  
  prev_slow = dal_df$slow_ma[t-1]
  prev_fast = dal_df$fast_ma[t-1]
  
  # use 'if' statement to translate crossover into signals
  if(current_fast > current_slow & prev_fast < prev_slow) {signal = "B"} else 
    if(current_fast < current_slow & prev_fast > prev_slow) {signal = "S"} else
      {signal = "H"}
  
  dal_df$signal[t+1] = signal
}

subset(dal_df[200:210,], select = c("Date","Close", "slow_ma", "fast_ma", "signal"))
```


## Holding Status and Investment Return

### APPLE Stock

After generating the signals, we must calculate the return from our strategy:

Store in `investment_return` vector

```{r, echo=FALSE}
# used to calculate fill price penalty / skid
# we will penalize by 40% from the open - high for buy orders
# and 40% from the open - low for sell orders

buy_weight = .35
sell_weight = -.35
```

```{r, echo=FALSE}
# Generate a holding status column  
# and investment return column

raw_data$holding = 0      # first record, you are not holding
raw_data$Inv_return = NA  # obviously the return will also be 0

# we can begin the loop
for (t in 100:nrow(raw_data)){
  
  if(t==1){prev_holding=0} 
  else {prev_holding=raw_data$holding[t-1]}
  
  # look at the signal to decide the change of holding status
  signal = raw_data$signal[t]
  if(signal=="B"){raw_data$holding[t]=1} else 
    if (signal=="S") {raw_data$holding[t]=0} else 
      if (signal=="H"){raw_data$holding[t]=prev_holding}
  
  # Now calculate investment return
  Open = raw_data$Open[t]
  Close = raw_data$Close[t]
  High = raw_data$High[t]
  Low = raw_data$Low[t]
  
  if(prev_holding==0 & signal=="B"){investment_return=Close/((High-Open)*buy_weight+Open) - 1} else 
    if(prev_holding==0 & signal=="H"){investment_return=0} else 
      #if(prev_holding==1 & signal=="S"){investment_return=Close[t-1]/(Open*(1+sell_weight)) - 1} else
      #if(prev_holding==0 & signal=="S"){investment_return=(-1)*(Close/(Open*(1+sell_weight)) - 1)} else
        if(prev_holding==1 & signal=="B"){investment_return=Close/raw_data$Close[t-1]-1} else
          if(prev_holding==1 & signal=="H"){investment_return=Close/raw_data$Close[t-1]-1} else
            {investment_return = Open/raw_data$Close[t-1]-1}
 
   # save the investment return to the dataset
  raw_data$Inv_return[t] = investment_return}
```

### DAL Stock

```{r, echo=FALSE}
# Generate a holding status column  
# and investment return column

dal_df$holding = 0      # first record, you are not holding
dal_df$Inv_return = NA  # obviously the return will also be 0

# we can begin the loop
for (t in 100:nrow(dal_df)){
  
  if(t==1){prev_holding=0} 
  else {prev_holding=dal_df$holding[t-1]}
  
  # look at the signal to decide the change of holding status
  signal = dal_df$signal[t]
  if(signal=="B"){dal_df$holding[t]=1} else 
    if (signal=="S") {dal_df$holding[t]=0} else 
      if (signal=="H"){dal_df$holding[t]=prev_holding}
  
  # Now calculate investment return
  Open = dal_df$Open[t]
  Close = dal_df$Close[t]
  High = dal_df$High[t]
  Low = dal_df$Low[t]
  
  if(prev_holding==0 & signal=="B"){investment_return=Close/((High-Open)*buy_weight+Open) - 1} else 
    if(prev_holding==0 & signal=="H"){investment_return=0} else 
      #if(prev_holding==1 & signal=="S"){investment_return=Close[t-1]/(Open*(1+sell_weight)) - 1} else
      #if(prev_holding==0 & signal=="S"){investment_return=(-1)*(Close/(Open*(1+sell_weight)) - 1)} else
        if(prev_holding==1 & signal=="B"){investment_return=Close/dal_df$Close[t-1]-1} else
          if(prev_holding==1 & signal=="H"){investment_return=Close/dal_df$Close[t-1]-1} else
            {investment_return = Open/dal_df$Close[t-1]-1}
 
   # save the investment return to the dataset
  dal_df$Inv_return[t] = investment_return}
```


## Cumulative Return

### Apple Stock

From the individual daily returns, we must create a cash and number of stocks vector to hold each days current figures.

Calculate and store in `cash` and `n_stock` vector:

```{r, echo=FALSE}
# starting investment $100,000
initial_wealth = 100000
# create vectors to store updates
raw_data$cash = rep(initial_wealth, length(raw_data$Close))
raw_data$n_stock = rep(0, length(raw_data$Close))

# loop through signals and simulate the trades
for (t in 1:length(raw_data$signal)){
  
  if(t==1){prev_cash = initial_wealth; n_stock = 0
    
    } else
      
      {prev_cash = raw_data$cash[t-1]; 
      n_stock = raw_data$n_stock[t-1]}
  
  signal = raw_data$signal[t]
  Open = raw_data$Open[t]
  Close = raw_data$Close[t]
  High = raw_data$High[t]
  Low = raw_data$Low[t]
  
# is.na to deal with rows at  beginning with no signal
  
  if(is.na(signal)==TRUE){
    prev_cash = raw_data$cash[t-1]
    n_stock = raw_data$n_stock[t-1] 
    
    } else
      
      if (signal=="B"){
        buy_skid = (High-Open) * buy_weight + Open
        trans_prc = buy_skid
        n_change = floor(prev_cash/trans_prc)
        stock_n = n_stock + n_change
        cash = prev_cash - n_change * trans_prc
            
        raw_data$cash[t] = cash
        raw_data$n_stock[t] = stock_n
        
        } else
          
          if (signal =="S"){
            sell_skid = (Open-Low) * sell_weight + Open
            trans_prc = sell_skid
            n_change = n_stock
            stock_n = n_change - n_stock
            cash = prev_cash + n_change * trans_prc
        
            raw_data$cash[t] = cash
            raw_data$n_stock[t] = stock_n
            
            }else
              
              if (signal == "H") {
                raw_data$cash[t] = raw_data$cash[t-1]
                raw_data$n_stock[t] = raw_data$n_stock[t-1]
                
                }
}
```

### DAL Stock

```{r, echo=FALSE}
# starting investment $100,000
initial_wealth = 100000
# create vectors to store updates
dal_df$cash = rep(initial_wealth, length(dal_df$Close))
dal_df$n_stock = rep(0, length(dal_df$Close))

# loop through signals and simulate the trades
for (t in 1:length(dal_df$signal)){
  
  if(t==1){prev_cash = initial_wealth; n_stock = 0
    
    } else
      
      {prev_cash = dal_df$cash[t-1]; 
      n_stock = dal_df$n_stock[t-1]}
  
  signal = dal_df$signal[t]
  Open = dal_df$Open[t]
  Close = dal_df$Close[t]
  High = dal_df$High[t]
  Low = dal_df$Low[t]
  
# is.na to deal with rows at  beginning with no signal
  
  if(is.na(signal)==TRUE){
    prev_cash = dal_df$cash[t-1]
    n_stock = dal_df$n_stock[t-1] 
    
    } else
      
      if (signal=="B"){
        buy_skid = (High-Open) * buy_weight + Open
        trans_prc = buy_skid
        n_change = floor(prev_cash/trans_prc)
        stock_n = n_stock + n_change
        cash = prev_cash - n_change * trans_prc
            
        dal_df$cash[t] = cash
        dal_df$n_stock[t] = stock_n
        
        } else
          
          if (signal =="S"){
            sell_skid = (Open-Low) * sell_weight + Open
            trans_prc = sell_skid
            n_change = n_stock
            stock_n = n_change - n_stock
            cash = prev_cash + n_change * trans_prc
        
            dal_df$cash[t] = cash
            dal_df$n_stock[t] = stock_n
            
            }else
              
              if (signal == "H") {
                dal_df$cash[t] = dal_df$cash[t-1]
                dal_df$n_stock[t] = dal_df$n_stock[t-1]
                
                }
}
```


## **Final Return**

### Apple Stock

Calculate the cash plus the value of the stock on hand for the last day in the sample.

Print final return from trend following system and buy-and-hold strategy:

```{r, echo=FALSE}
## Trend-following system
# Calculations for final portfolio value
final_return = raw_data$n_stock[max(index(raw_data))] * raw_data$Close[max(index(raw_data))] + raw_data$cash[max(index(raw_data))-1]

print(paste("Initial Investment:", sprintf("$ %3.2f", initial_wealth)))
print(paste("Trend-following system Final Return:",sprintf("$ %3.2f" , final_return)))


## Buy and hold strategy
initial_price = raw_data$Open[1]
final_price = raw_data$Close[max(index(raw_data))]

n_purchased = floor(initial_wealth/initial_price)

total_return = (final_price - initial_price) * n_purchased
print(paste("Buy and hold system Final Return:",sprintf("$ %3.2f" , total_return)))
```

### Delta Airlines Stock
```{r, echo=FALSE}
## Trend-following system
# Calculations for final portfolio value
final_return_dal = dal_df$n_stock[max(index(dal_df))] * dal_df$Close[max(index(dal_df))] + dal_df$cash[max(index(dal_df))-1]

print(paste("Initial Investment:", sprintf("$ %3.2f", initial_wealth)))
print(paste("Trend-following system Final Return:",sprintf("$ %3.2f" , final_return_dal)))


## Buy and hold strategy
initial_price_dal = dal_df$Open[1]
final_price_dal = dal_df$Close[max(index(dal_df))]

n_purchased_dal = floor(initial_wealth/initial_price_dal)

total_return_dal = (final_price_dal - initial_price_dal) * n_purchased_dal
print(paste("Buy and hold system Final Return:",sprintf("$ %3.2f" , total_return_dal)))
```

# Visualize Days with Largest Loss and Gain

```{r, echo=FALSE}
max_gain=raw_data[which.max(raw_data$Inv_return),]
max_loss = raw_data[which.min(raw_data$Inv_return),]

# max(raw_data$Inv_return, na.rm = TRUE)

# Create subset
max_data = raw_data[3310:3330,]
min_data = raw_data[3310:3330,]
max_loss
max_gain
```



```{r, echo=FALSE}
# Create arrow annotations

# Upper annotations
h_a <- list(
  x = max_gain$Date,
  y = max_gain$Price,
  text = "Largest Gain",
  xref = "x",
  yref = "y",
  showarrow = TRUE,
  arrowcolor = "green",
  arrowhead = 5,
  arrowsize = 0.8,
  arrowwidth = 2.5,
  opacity = 0.75,
  align = "right",
  ax = 5,
  ay = -55
)

# Lower annotations

l_a <- list(
  x = max_loss$Date,
  y = max_loss$Price,
  text = "Largest Loss",
  xref = "x",
  yref = "y",
  showarrow = TRUE,
  arrowcolor = "green",
  arrowhead = 5,
  arrowsize = 0.8,
  arrowwidth = 2.5,
  opacity = 0.75,
  align = "right",
  ax = -5,
  ay = 55
)


# figure labels
f <- list(
  family = "Courier New, monospace",
  size = 18,
  color = "#7f7f7f ")
x <- list(
  title = "x Axis",
  titlefont = f)
y <- list(
  title = "y Axis",
  titlefont = f)
```

## Largest Gain

```{r, echo=FALSE}
fig <- max_data %>% plot_ly(x = ~Date, type="ohlc",
          open = ~Open, close = ~Close,
          high = ~High, low = ~Low, name = "AAPL",
          increasing = i, decreasing = d)

# color of bars for chart
i <- list(line = list(color = '#17BECF')) # 'i' for increasing
d <- list(line = list(color = '#b22222')) # 'd' for decreasing

# Add Fast and Slow moving average lines
fig <- fig %>% add_lines(x = ~Date, y = ~slow_ma, name = "Slow",
            line = list(color = '#ccc', width = 0.5),
            legendgroup = "Bands", inherit = F,
            showlegend = TRUE, hoverinfo = "none") 
fig <- fig %>% add_lines(x = ~Date, y = ~fast_ma, name = "Fast",
            line = list(color = '#E377C2', width = 0.5),
            hoverinfo = "none", inherit = F)

# Add y-axis title
fig <- fig %>% layout(yaxis = list(title = "Price"))

# Add arrow annotations
fig <- fig %>% layout(annotations = h_a)

# Add title
fig <- fig %>% layout(title = "AAPL Stock - Strategy Largest 1-Day Gain",
                       xaxis = list(rangeslider = list(visible = F)))

fig
```

## Largest Loss

```{r, echo=FALSE}
fig <- min_data %>% plot_ly(x = ~Date, type="ohlc",
          open = ~Open, close = ~Close,
          high = ~High, low = ~Low, name = "AAPL",
          increasing = i, decreasing = d)

# color of bars for chart
i <- list(line = list(color = '#17BECF')) # 'i' for increasing
d <- list(line = list(color = '#b22222')) # 'd' for decreasing

# Add Fast and Slow moving average lines
fig <- fig %>% add_lines(x = ~Date, y = ~slow_ma, name = "Slow",
            line = list(color = '#ccc', width = 0.5),
            legendgroup = "Bands", inherit = F,
            showlegend = TRUE, hoverinfo = "none") 
fig <- fig %>% add_lines(x = ~Date, y = ~fast_ma, name = "Fast",
            line = list(color = '#E377C2', width = 0.5),
            hoverinfo = "none", inherit = F)

# Add y-axis title
fig <- fig %>% layout(yaxis = list(title = "Price"))

# Add arrow annotations
fig <- fig %>% layout(annotations = l_a)

# Add title
fig <- fig %>% layout(title = "AAPL Stock - Strategy Largest 1-day Loss",
                       xaxis = list(rangeslider = list(visible = F)))

fig
```

## Max Adverse & Favorable Single-Day Returns

-   Retrieve the days with the largest single-day loss, gain and average daily return for a simple buy and hold system, as well as for our trend-following system.

-   Compare the results from the two strategies.

```{r, echo=FALSE}
max_daily_return = round((max(raw_data$Inv_return, na.rm = TRUE)*100),4)
max_daily_loss =  round((min(raw_data$Inv_return, na.rm = TRUE)*100), 4)
avg_daily_return = round((mean(raw_data$Inv_return, na.rm = TRUE)*100),4)

bh_max_return = round((max(raw_data$return, na.rm = TRUE)*100),4)
bh_max_loss =  round((min(raw_data$return, na.rm = TRUE)*100), 4)
bh_avg_return = round((mean(raw_data$return, na.rm = TRUE)*100),4)

print(paste("Largest 1-day return with trend system:", max_daily_return,"%"))
print(paste("Largest 1-day loss with trend system:", max_daily_loss,"%"))
print(paste("Average 1-day with trend system return:", avg_daily_return,"%"))

print(paste("Largest 1-day return buy and hold strategy:", bh_max_return,"%"))
print(paste("Largest 1-day loss with buy and hold strategy:", bh_max_loss,"%"))
print(paste("Average 1-day with buy and hold strategy:", bh_avg_return,"%"))
```

# Performance Evaluation

## Distribution of Returns

### Apple Stock

Create a histogram of returns for a simple buy and hold strategy, as well as for our trend-following system:

```{r echo=FALSE}
p <- raw_data %>%
  ggplot(aes(x=Inv_return)) +
    geom_histogram(fill="#69b3a2", color="#e9ecef", alpha=0.7, binwidth = .005) +
  coord_cartesian(xlim = c(-0.07, 0.07), ylim = c(0, 500)) +
  labs(x = "Investment Returns", y = "Frequency of Returns",
       title = "AAPL EMA Crossover - Histogram of Investment Returns",
       caption = "Trend following system - Slow lag: 150 periods, Fast lag: 25 periods")+
  theme_classic()
p


bh = raw_data %>%
  ggplot(aes(x=return)) +
    geom_histogram(fill="#69b3a2", color="#e9ecef", alpha=0.7, binwidth = .005) +
  coord_cartesian(xlim = c(-0.07, 0.07), ylim = c(0, 500)) +
  labs(x = "Investment Returns", y = "Frequency of Returns",
       title = "AAPL Buy / Hold - Histogram of Investment Returns",
       caption = "This histogram considers a buy and hold strategy")+
  theme_classic()
bh

ggsave("aapl_returns_hist.png", plot = p)
ggsave("aapl_buy_hold_returns.png", plot = bh)
```

### Delta Airlines Stock

```{r, echo=FALSE}
p_dal <- dal_df %>%
  ggplot(aes(x=Inv_return)) +
    geom_histogram(fill="#69b3a2", color="#e9ecef", alpha=0.7, binwidth = .005) +
  coord_cartesian(xlim = c(-0.07, 0.07), ylim = c(0, 280)) +
  labs(x = "Investment Returns", y = "Frequency of Returns",
       title = "DAL EMA Crossover - Histogram of Investment Returns",
       caption = "Trend following system - Slow lag: 150 periods, Fast lag: 25 periods")+
  theme_classic()
p_dal

ggsave("dal_returns_hist.png", plot = p_dal)
```
## Mean and Standard Deviation

### AAPL

Retrieve the average daily return and volatility of our trend-following system, as well as a simple buy and hold strategy:

```{r, echo=FALSE}
average_return = mean(raw_data$Inv_return, na.rm = TRUE) # average return
volatility = sd(raw_data$Inv_return, na.rm = TRUE) # volatility

print(paste("Average return:", round(average_return,6)))
print(paste("Volatility:", round(volatility,6)))


bh_average_return = mean(raw_data$return, na.rm = TRUE) # average return
bh_volatility = sd(raw_data$return, na.rm = TRUE) # volatility

print(paste("Average return buy hold system:", round(bh_average_return,6)))
print(paste("Volatility buy hold system:", round(bh_volatility,6)))
print(paste("Buy and hold strategy experiences larger volatility for a small amount of additional return"))
```

### DAL
```{r, echo=FALSE}
average_return_dal = mean(dal_df$Inv_return, na.rm = TRUE) # average return
volatility_dal = sd(dal_df$Inv_return, na.rm = TRUE) # volatility

print(paste("DAL Average return:", round(average_return_dal,6)))
print(paste("DAL Volatility:", round(volatility_dal,6)))
print(paste("As we can see, the average return when using this system for DAL is much smaller and the volatility is larger. We can expect the Sharpe ratio to be much smaller in this environment"))
```
## Sharpe Ratio

### Apple Stock

The Sharpe ratio was developed by Nobel laureate [William F. Sharpe](https://www.investopedia.com/terms/w/william-f-sharpe.asp) and is used to help [investors](https://www.investopedia.com/terms/i/investor.asp) understand the [return of an investment](https://www.investopedia.com/terms/r/returnoninvestment.asp) compared to its risk.12 The ratio is the average return earned in excess of the risk-free rate per unit of [volatility](https://www.investopedia.com/terms/v/volatility.asp) or total risk. Volatility is a measure of the price fluctuations of an asset or portfolio.

-   Our **Sharpe** ratios are based on **daily returns** for both the buy and hold strategy and the trend-following system.

```{r, echo=FALSE}
rf = .00001 # risk-free rate
print(paste("AAPL Trend following: Sharpe ratio of daily returns:",round((average_return-rf)/volatility,3)))
print(paste("AAPL Buy & hold: Sharpe ratio of daily returns:",round((bh_average_return-rf)/bh_volatility,3)))

print(paste("Our AAPL Trend-following system achieves a slightly higher Sharpe ratio"))
```

### Delta Stock

```{r, echo=FALSE}
rf = .00001 # risk-free rate
print(paste("DAL Trend following: Sharpe ratio of daily returns:",round((average_return_dal-rf)/volatility_dal,5)))

print(paste("Compared to Apple Trend-following returns, DAL's example provides a much smaller Sharpe ratio - this is due to whipsaw-like nature of Delta's historical stock price and the trend-following's lack of ability to trade in this type of environment (at least with the parameters we have chosen)"))
```

## Value at Risk

### Apple Stock

**VAR** measures the cut-off return that your financial asset will fall below with certain probability.

Value at risk (VaR) is a statistic that quantifies the extent of possible financial losses within a firm, portfolio, or position over a specific time frame. This metric is most commonly used by [investment ](https://www.investopedia.com/terms/i/investmentbank.asp)and [commercial banks](https://www.investopedia.com/terms/c/commercialbank.asp) to determine the extent and probabilities of potential losses in their institutional portfolios.

```{r, echo=FALSE}
var = quantile(raw_data$Inv_return, 0.05, na.rm = TRUE)*100
bh_var = quantile(raw_data$return, 0.05, na.rm = TRUE)*100

print(paste("Apple Trend-following VAR: we can expect 95% of the time returns will be greater than:", round(var,4)))
print(paste("Apple Buy / Hold model VAR: we can expect 95% of the time returns will be greater than:", round(bh_var,4)))
```

### Delta Airlines

```{r, echo=FALSE}
var_dal = quantile(dal_df$Inv_return, 0.05, na.rm = TRUE)*100


print(paste("DAL Trend-following VAR: we can expect 95% of the time returns will be greater than:", round(var_dal,4)))
```

# Summary of Investment & Market Return

In order to break down risk into two components, we need to collect what part of the return was attributed to the market and what part to our trading system.

## Fit CAPM Model

-   Our $y$ is the excess return of the trading system

-   Our $x$ is the excess return of the market

```{r, echo=FALSE}
y = raw_data$Inv_return - rf
x = SPX$return - rf
```

```{r, echo=FALSE}
# fit regression

capm = lm(y~x)
summary(capm)

print(paste("Trend-following system Beta:", round(capm$coefficients[2],4)))
```

The larger the $BETA$ is, the larger the systematic risk is.

We use this to measure the aggressiveness of the stock or the trading system.

Our trend-following system is mostly conservative.

## Jensen's Alpha (Abnormal Return)


-   The regression model's intercept term tells us how much the strategy gives us extra reward, above the market return.

-   Generally we would like to see a positive $ALPHA$

    -   This is associated with its **statistical significance**.

    -   The higher $ALPHA$ the better the investment performance

```{r, echo=FALSE}
tf_beta = capm$coefficients[2] # beta
jensens_alpha = capm$coefficients[1] # Jensen's Alpha

print(paste("Jensen's Alpha:", round(jensens_alpha,4)))
```

## Treynor's Ratio

```{r, echo=FALSE}
print(paste("Treynor's Ratio Version 1:", round(jensens_alpha/tf_beta, 4)))
```
```{r, echo=FALSE}
print(paste("Treynor's Ratio Version 2:", round(((mean(raw_data$Inv_return, na.rm = TRUE))/tf_beta),4)))
```


# Historical Portfolio Return

```{r}
raw_data$portfolio_value = raw_data$n_stock*raw_data$Close + raw_data$cash
```

```{r, echo=FALSE}
# Create arrow annotations for trade entry signal

# AAPL
# Upper annotations
h_a <- list(
  x = h$Date,
  y = h$slow_ma,
  text = "Sell",
  xref = "x",
  yref = "y",
  showarrow = TRUE,
  arrowcolor = "red",
  arrowhead = 5,
  arrowsize = 1,
  arrowwidth = 1.7,
  opacity = 0.75,
  align = "left",
  ax = 5,
  ay = -55
)

# Lower annotations

l_a <- list(
  x = l$Date,
  y = l$slow_ma,
  text = "Buy",
  xref = "x",
  yref = "y",
  showarrow = TRUE,
  arrowcolor = "green",
  arrowhead = 5,
  arrowsize = 1,
  arrowwidth = 1.7,
  opacity = 0.75,
  align = "right",
  ax = -5,
  ay = 55
)


# figure labels
f <- list(
  family = "Courier New, monospace",
  size = 24,
  color = "#7f7f7f ")
x <- list(
  title = "x Axis",
  titlefont = f)
y <- list(
  title = "y Axis",
  titlefont = f)
```

```{r}
# Plot main data
fig <- raw_data %>% plot_ly(x = ~Date, type="ohlc",
          open = ~Open, close = ~Close,
          high = ~High, low = ~Low, name = "AAPL",
          increasing = i, decreasing = d)

# Add Fast and Slow moving average lines
fig <- fig %>% add_lines(x = ~Date, y = ~slow_ma, name = "Slow EMA",
            line = list(color = '#ccc', width = 0.75),
            legendgroup = "Bands", inherit = F,
            showlegend = TRUE, hoverinfo = "none") 
fig <- fig %>% add_lines(x = ~Date, y = ~fast_ma, name = "Fast EMA",
            line = list(color = '#E377C2', width = 0.75),
            legendgroup = "bands",
            hoverinfo = "none", inherit = F)

ay <- list(
  tickfont = list(color = "grey"),
  overlaying = "y",
  side = "right",
  title = "<b>Secondary:</b> Portfolio Value")

fig <- fig %>% add_lines(x = ~Date, y = ~portfolio_value, name = "Portfolio Value",
            line = list(color =  'limegreen', width = 1), dash = 'dot', yaxis = "y2")

# Add y-axis title
fig <- fig %>% layout(yaxis = list(title = "Price"))


# Add arrow annotations
fig <- fig %>% layout(annotations = h_a)
fig <- fig %>% layout(annotations = l_a)



# Add title
fig <- fig %>% layout(
  title = "<b>Trend-Following System:</b> Portfolio Value", yaxis2 = ay,
  xaxis = list(rangeslider = list(visible = F)),
  yaxis = list(title = "<b>Primary:</b> Stock Price")
  ) %>%
  layout(plot_bgcolor = 'rgb(255, 255, 255)',
         xaxis = list(
            zerolinecolor = '#ffffff',
            zerolinewidth = 2,
            gridcolor = '#ffffff'),
          yaxis = list(
            zerolinecolor = '#ffffff',
            zerolinewidth = 2,
            gridcolor = '#ffffff')
          )


fig
```



















